---
title: "Comanche County 00-19"
author:
  - Leticia, Fellow
  - Open Justice Oklahoma
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
library(ojo)
library(knitr)


nc <- function(x) {comma(x, digits = 0)}

```

Background information: Why are we researching this topic? What is the expected product? A blog post? Analysis for an internal or external partner?

Using court data, this project aims to shed new light on the extent of the medical debt problem in Oklahoma.

> **Research Question**

# Data sources

Where is the data from? The ojo database? OSCN? ODCR? Crime in Oklahoma reports?

The data is from the ojo database and OSCN.  

## Timeframe

What years/months are you looking at? Year of case filing, first disposition, prison admission, etc.? Be as specific as possible.

The analysis aims to identify trends in medical debt cases over the past 20 years (2000-2019).

## Geography

What courts/counties/states are you looking at?

This project will focus on medical debt collection cases in civil courts across Oklahoma.


## Variables

What variables did you use? What types of cases, crimes, etc.?

Debt collections in the state's District Court.


## Query

If you're using data from the OJO database, include the query you used to pull your data. This may be a function that starts `ojo_query_` or `dbGetQuery`.

``` {r}

sc <- ojo_query_disps("COMANCHE", "SC", 2000:2019) 



```

# Checking Data

Explain the results of the checks of your data.

The data is 99-100% complete. 

## Check for Completeness

If you're using court data, you can use the `ojo_check_comp` to see how complete the data is.

```{r}

ojo_check_comp(sc)

completeness
```

## Check for NAs

Detail the NAs and blank items in your data and, if necessary and possible, fill them in.

A large portion of the "iss_plaint" data for the early 2000s is missing. 

``` {r}

nas_plaint<- sc %>%
  group_by(casenum) %>% 
  filter(all(is.na(iss_plaint) | iss_plaint == "")) 

nas_plaint

nas_plaint <- nas_plaint %>% 
  group_by(casenum) %>% 
  slice(1)

year_nas_plaint <- nas_plaint%>% 
  ungroup %>%
  count(file_year)

year_nas_plaint



```
``` {r}

nas_desc <- sc %>%
  group_by(casenum) %>% 
  filter(all(is.na(iss_desc) | iss_desc == "")) 

nas_desc

nas_desc <- nas_desc%>% 
  group_by(casenum) %>% 
  slice(1)

year_nas_desc <- nas_desc%>% 
  ungroup %>%
  count(file_year)

year_nas_desc

```


# Defining Measures

This is where you'll likely spend the biggest chunk of time if you're using court data. There are a lot of ways things can go wrong, so we want to be able to trace back all of our steps and document all the decisions we're making.

Because each analysis project is a bit different, the number and length of subsections in this section will vary a lot. 


``` {r}
sc %>% 
  count(iss_desc) %>% 
  arrange(desc(n)) 
```
``` {r}
debt <- sc %>% 
  filter(!str_detect(iss_desc, "FORC|REPLEVIN|BREACH"))

debt %>% 
  count(iss_desc) %>% 
  arrange(desc(n)) 
```
``` {r}
debt <- debt %>% 
  filter(str_detect(iss_desc, "DEBT"))

debt %>% 
  count(iss_desc) %>% 
  arrange(desc(n)) 
```
``` {r}
debt %>% 
  count(iss_plaint) %>% 
  arrange(desc(n))  
```
``` {r}
debt <- debt %>% 
  mutate(plaint_type = case_when(str_detect(iss_plaint, "HEALTH|CLINIC|MEDICAL|HOSPITAL|IMAGING\\sCENTER|RADIOLOGY|PEDIATRICS|SURGERY|ALLERGY\\sEAR\\sNOSE\\s&\\sTHROAT|HOPSITAL|DOCTORS\\sMRI|HOSPTIAL|NOSE\\s&\\sTHROAT|EAR\\sNOSE|THROAT\\sINSTITUTE") ~ "MEDICAL",
                                 str_detect(iss_plaint, "BANK|FCU|CREDIT|LOAN|FINANC|CASH|MONEY|WORLD\\sACCEPTANCE|FURNITURE|FUNDING|REALTY|CABLEVISION|ELECTRIC|PAYDAY|PAY\\sDAY|ADVANCE\\sAMERICA|BANCFIRST|BOND|RENTALS|MACARTHUR\\sPARK\\sHOMES|FSNB|CHECKMATE|CARMART|CAR\\sMART|SYCAMORE\\sTRAIL|WESTERN\\sSHAMROCK|AUTO|LEASING|PLUMBING|MECHANICA|APARTMENTS|CHECKSMART|DREAM\\sHOMES|FIDELITY|HOUSING\\sAUTHORITY|PIPPIN\\sBROTHERS|CAR|KLEIN\\sGROUP|HOME\\sFINDERS|REAL\\sESTATE|WATER\\sCONDITIONING") ~ "OTHER",
                                 TRUE ~ "UNKNOWN"))

debt %>% 
  count(plaint_type) %>% 
  arrange(desc(n))


```
``` {r}
debt %>%
  filter(plaint_type == "UNKNOWN") %>% 
  count(iss_plaint) %>% 
  arrange(desc(n)) 
```

```{r}
debt %>% 
  count(plaint_type) %>% 
  arrange(desc(n))
```
```{r}
debt %>%
  filter(plaint_type == "MEDICAL") %>% 
  count(iss_plaint) %>% 
  arrange(desc(n)) 

```
``` {r}
debt <- debt %>% 
  filter(plaint_type == "MEDICAL")

```

## Measure 1

Describe how you defined the first measure.

# Summarizing and Visualizing Data

There will likely be several sections here too. Summarize and visualize the data in different ways and identify trends that you see.
``` {r}
debt <- debt %>% 
  group_by(casenum) %>% 
  slice(1)

year_sum <- debt %>% 
  ungroup %>%
  count(file_year)

ggplot(year_sum, aes(file_year, n)) +
  geom_line() +
  geom_text(aes(y = n + 50, label = n), family = "Menlo", size = 3) +
  xlab("Year") + ylab("Cases") +
  ylim(0, NA) +
  theme_ojo() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.subtitle = element_text(hjust = 0.5)) +
  labs(title = "Small claims medical debt cases",
       subtitle = "Comanche County 2000-2019")
```
``` {r}
debt <- debt %>% 
  mutate(plaint_cat = case_when(str_detect(iss_plaint, "COMANCHE COUNTY MEMORIAL") ~ "COMANCHE MEMORIAL HOSPITAL",
                                str_detect(iss_plaint, "THE COMANCHE COUNTY HOSPITAL AUTHORITY") ~ "COMANCHE HOSPITAL AUTHORITY",
                                str_detect(iss_plaint, "COMANCHE COUNTY HEALTHCARE") ~ "COMANCHE HEALTHCARE",
                                TRUE ~ "OTHER"))

plaint_sum <- debt %>% 
  ungroup %>% # Our data frame is still grouped by casenum from the code above; ungroup to summarize using broader groups
  count(file_year, plaint_cat)

ggplot(plaint_sum, aes(file_year, n, group = plaint_cat, color = plaint_cat)) +
  geom_line() +
  xlab("Year") + ylab("Cases")+
  ylim(0, NA) + # Extends y-axis down to zero
  theme_ojo() + # Adds ojo styling to the plot
  theme(legend.title=element_blank())+
  theme(plot.title = element_text(hjust = 0.5, size = 14)) +
  theme(plot.subtitle = element_text(hjust = 0.5, size = 13)) +
  geom_text(aes(y = n + 50, label = n), family = "Menlo", size = 3) +
  theme(legend.text=element_text(size = 7.5))+
  theme(legend.position="bottom") +
  labs(title = "Small claims medical debt cases\nin Comanche County 2000-2019",
       subtitle = "By medical provider") +
  scale_color_manual(values = ojo_pal) # Gives lines colors from ojo's palette



```


## Summary 1

- How you summarized and why
- How you visualized the data
- Potential leads and trends

```{r results="asis", echo=FALSE}
cat("
<style>
body {
  padding-top: 63px;
}

h1.title.toc-ignore {
  font-family: 'Pluto Sans';
  font-weight: bold;
  background-color: #F8D64E;
}

h1, h2, h3 {
  font-family: 'Pluto Sans';
  font-weight: bold;
}

#TOC {
  font-family: 'Menlo'
}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    font-weight: bold;
    color: black;
    background-color: #F8D64E;
}

p a {
  color: black;
  background-color: #F8D64E;
  font-weight: bold;
}

</style>
")
```
